Approximately Exact Calculations for Linear Mixed Models

### 2015-06-08 ms.tex
minor correction from Hodges

\appendix
\section{Appendix: Derivation of $\{a_j\}$ and $\{\hat v_j\}$ in \eqref{eq:reexpress}, \eqref{eq:logpost}, and \eqref{eq:sum}}
Our derivation follows \cite{hodges:2013},
which contains more details.  There are three steps.
\begin{enumerate}
  \item \textbf{Make the covariance matrices proportional to the identity.}  If $\Sigma_e$ is not the identity matrix,
    transform the data to $\Sigma_e^{-.5}y$.  The transformed data, which we shall still call $y$, has covariance
    proportional to the identity.  Similarly, if $\Sigma_s$ is not the identity matrix, re-parameterize the random
    effects to $\Sigma_s^{-.5}u$.  The re-parameterized random effects, which we shall still call $u$, have
    covariance proportional to the identity.
  \item \textbf{If the column spaces of $X$ and $Z$ have a non-trivial intersection, transform them.}
    Let $s_X=\text{rank}(X)$  and $s_Z=\text{rank}(X|Z) - s_X$.  Let $\Gamma_X$ be an $n \times s_X$ matrix whose
    columns are an orthonormal basis for the column space of $X$.  Let $\Gamma_Z$ be an $n \times s_Z$ matrix
    such that the columns of $\left[\Gamma_X | \Gamma_Z \right]$ are an orthonormal basis for the column space
    of $\left[X|Z\right]$.  Let $\Gamma_c$ be an $n \times (n-s_X-s_Z)$ matrix such that the columns of
    $\left[\Gamma_X | \Gamma_Z | \Gamma_c\right]$ are an orthonormal basis for $\mathbb{R}^n$.  Define the matrix
    \begin{equation*}
      M = \begin{bmatrix}
               M_{XX} & M_{XZ}\\
               0 & M_{ZZ}
             \end{bmatrix}
    \end{equation*}
    by $[X|Z] = [\Gamma_X|\Gamma_Z]M$ where $M_{XX}$ is $s_X \times p$ and $M_{XZ}$ is $s_Z \times q$.
    $\Gamma_X$ and $\Gamma_Z$ are transformed versions of $X$ and $Z$ that have non-overlapping
    column spaces.
  \item \textbf{Re-parameterize and diagonalize.}  Let $M_{ZZ}$ have the singular value decomposition
    $PA^{.5}L^t$.  Now the linear mixed model \eqref{eq:lmm} can be written as
    \begin{equation*}
      \begin{split}
        y &= \begin{bmatrix} X | Z \end{bmatrix} \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
        &= \begin{bmatrix} \Gamma_X | \Gamma_Z \end{bmatrix} M \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
        &= \begin{bmatrix} \Gamma_X | \Gamma_Z | P \end{bmatrix} \begin{bmatrix} \beta^* \\ v \end{bmatrix} + \epsilon
      \end{split}
    \end{equation*}
    where $\beta^* = M_{XX}\beta + M_{XZ}u$ and $v = A^{.5}L^t u$.  $\beta^*$ contains the re-parametrized
    fixed effects while $v$ contains the re-parametrized random effects.  The corresponding design matrices
    $\Gamma_X$ and $\Gamma_Z P$ are orthogonal to each other.
\end{enumerate}
Finally, the $\{a_j\}$ in \eqref{eq:reexpress} are the diagonal elements of $A$, all of which are strictly positive, and the $\{\hat v_j\}$ in \eqref{eq:reexpress} are given by $\hat v = (\hat v_1, \dots, \hat v_{s_Z})^t = P^t \Gamma_Z^t y$.

>>>>

\appendix
\section{Appendix: Derivation of $\{a_j\}$ and $\{\hat v_j\}$ in \eqref{eq:reexpress}, \eqref{eq:logpost}, and \eqref{eq:sum}}
Our derivation follows \cite{hodges:2013},
which contains more details.  There are three steps.
\begin{enumerate}
  \item \textbf{Make the covariance matrices proportional to the identity.}  If $\Sigma_e$ is not the identity matrix,
    transform the data to $\Sigma_e^{-.5}y$.  The transformed data, which we shall still call $y$, has covariance
    proportional to the identity.  Similarly, if $\Sigma_s$ is not the identity matrix, re-parameterize the random
    effects to $\Sigma_s^{-.5}u$.  The re-parameterized random effects, which we shall still call $u$, have
    covariance proportional to the identity.
  \item \textbf{If the column spaces of $X$ and $Z$ have a non-trivial intersection, transform them.}
    Let $s_X=\text{rank}(X)$  and $s_Z=\text{rank}(X|Z) - s_X$.  Let $\Gamma_X$ be an $n \times s_X$ matrix whose
    columns are an orthonormal basis for the column space of $X$.  Let $\Gamma_Z$ be an $n \times s_Z$ matrix
    such that the columns of $\left[\Gamma_X | \Gamma_Z \right]$ are an orthonormal basis for the column space
    of $\left[X|Z\right]$.  Let $\Gamma_c$ be an $n \times (n-s_X-s_Z)$ matrix such that the columns of
    $\left[\Gamma_X | \Gamma_Z | \Gamma_c\right]$ are an orthonormal basis for $\mathbb{R}^n$.  Define the matrix
    \begin{equation*}
      M = \begin{bmatrix}
               M_{XX} & M_{XZ}\\
               0 & M_{ZZ}
             \end{bmatrix}
    \end{equation*}
    by $[X|Z] = [\Gamma_X|\Gamma_Z]M$ where $M_{XX}$ is $s_X \times p$ and $M_{XZ}$ is $s_Z \times q$.
    $\Gamma_X$ and $\Gamma_Z$ are transformed versions of $X$ and $Z$ that have non-overlapping
    column spaces.
  \item \textbf{Re-parameterize and diagonalize.}  Let $M_{ZZ}$ have the singular value decomposition
    $PA^{.5}L^t$.  Now the linear mixed model \eqref{eq:lmm} can be written as
    \begin{equation*}
      \begin{split}
        y &= \begin{bmatrix} X | Z \end{bmatrix} \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
        &= \begin{bmatrix} \Gamma_X | \Gamma_Z \end{bmatrix} M \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
        &= \begin{bmatrix} \Gamma_X | \Gamma_Z P \end{bmatrix} \begin{bmatrix} \beta^* \\ v \end{bmatrix} + \epsilon
      \end{split}
    \end{equation*}
    where $\beta^* = M_{XX}\beta + M_{XZ}u$ and $v = A^{.5}L^t u$.  $\beta^*$ contains the re-parametrized
    fixed effects while $v$ contains the re-parametrized random effects.  The corresponding design matrices
    $\Gamma_X$ and $\Gamma_Z P$ are orthogonal to each other.
\end{enumerate}
Finally, the $\{a_j\}$ in \eqref{eq:reexpress} are the diagonal elements of $A$, all of which are strictly positive, and the $\{\hat v_j\}$ in \eqref{eq:reexpress} are given by $\hat v = (\hat v_1, \dots, \hat v_{s_Z})^t = P^t \Gamma_Z^t y$.
